Penalized linear mixed models for structured genetic data

### 2021-03-04 main.tex
Additional edits throughout - reorganizing, making more 'friendly'

An additional hurdle such analyses must contend with is that of population structure such as population stratification or cryptic relatedness. The presence of such structure can adversely affect analyses in several ways. Firstly, allele frequency differences due to ancestry within a sample can be mistaken for biological signal if steps are not taken to mitigate this effect. One common solution, chopping up a sample and performing ancestry-specific analyses, often results in a loss of statistical power from reduced sample sizes. Secondly, correlation among subjects within a sample, whether due to common ancestry or more recent relatedness, violates the independent errors assumption that many regression methods rely on, rendering statistical methods such as linear and logistic invalid.

@@

As with multiple linear regression models, lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, the majority of this work has focused on plant and animal breeding data, a setting which affords much greater experimental control than studies of human genetics \citep{Amin2007, hoffman2013correcting, price2006principal, rakitsch2013lasso, bhatnagar2020simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principle be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and confounding arising from observational study \citep{lawson2019population, barton2019population}.

@@

\begin{equation}
    Q(\bbeta | \lam) = \frac{1}{2n}|| \by -  \beta_0 - \bX \bbeta ||_2^2 + \lam ||\bbeta||_1,
\end{equation}
where $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||\mathbf{x}||_2 = \sqrt{\sum_{i=1}^p x_i^2}$ the $\ell_2$ norm. The first component of the objective function is the linear regression loss function, which quantifies the overall model fit and captures how well the estimated parameters agree with the observed data. The second component is the $\ell_1$ penalty applied to the estimated coefficient values (but not the intercept, $\beta_0$), which discourages large, unrealistic parameter values. The tuning parameter, $\lambda$, controls the tradeoff between these two components. For sufficiently large values of $\lambda$ all elements of $\bbeta$ will be estimated to be 0. In the $p \ll n$ setting, if $\lam = 0$, the lasso solution coincides with that of ordinary least squares (OLS). The lasso assumes observations are independent. However, the $\ell_1$ penalty of the lasso can also be applied to nonlinear models such as Cox regression \citep{tibshirani1997lasso} and logistic regression \citep{shevade2003simple} by replacing the residual sum of squares with the appropriate negative log-likelihood function.

@@

A central issue explored in this work has been the potential for population stratification and environmental confounding to lead to the identification of spurious associations. In Section~\ref{sec:sel_bias} we characterized the three lasso methods (ordinary, PC and LMM) in terms of their susceptibility to this type of selection bias in a simplified setting of two subpopulations. We now seek to examine these methods' performance in a genome-wide, pathwise manner.

Figure \ref{fig:gw_pc1_corr} displays the mean absolute correlation of the non-causal SNPs selected for a particular $\lambda$ value with the first principal component (PC1) of the Empirical data for each method across the full $\lambda$ path. The corresponding shaded regions indicate the SE about the mean; overall these bounds are largest early on in each $\lambda$ path, where only a few SNPs are be selected, but narrow as $\lambda$ increases and more variables are allowed to enter the model.\anna{Do I need to explain why we are using correlation with the first PC as a proxy for MAF difference or is this obvious?} The absolute correlation with PC1 is used as a measure of how population-informative the selected SNPs are given the complex Empirical data structure compared to that in Section~\ref{sec:sel_bias}.

@@

\anna{The results of a genome-wide analysis of Empirical data agree with the findings of our simulation experiments, in particular that PC-lasso and LMM-lasso are able to substantially reduce selection bias compared to the ordinary lasso. Although PC-lasso was found to control this selection bias most effectively, it should be noted that because the Empirical data consists of three distinct ethnic groups, its RRM (Figure~\ref{fig:empirical}) demonstrates predominantly coarse population structure. Given our results in Section~\ref{sec:sim-fine-coarse}, we might expect LMM-lasso to provide greater improvement over the ordinary lasso in the analysis of more finely structured genome-wide data.}

\anna{The nature of the simulations and analyses in this work have allowed us to select penalty parameters based on the true number of causal features in a model, which allows for fair comparison across methods. In a real data scenario, the true number of causal SNPs would not be known a priori. As such, appropriate methods for penalty parameter selection in penalized LMMs are worth additional discussion. As of yet there is no clear consensus on the optimal way to select a penalty parameter for penalized LMMs.}

\textbf{\anna{Add reference to that paper that mentions this challenge explicitly}}

\anna{Several authors have used the Bayesian information criterion (BIC) for variable selection in mixed models \cite{bondell2010joint, schelldorfer2011estimation}. However it has been shown that the Akaike information criterion (AIC) and BIC may perform poorly in the high-dimensional setting \citep{fan2013tuning}. The high-dimensional BIC introduced by \citet{fan2013tuning} penalizes model parameters more severely in an attempt to perform better in high dimensional settings. \citet{bhatnagar2020simultaneous} employs this high-dimensional BIC as its default method to select $\lambda$.}

\anna{Subsampling approaches, such as CV and stability selection, are another common class of variable selection methods often preferred to information criterion-based approaches. \citet{rakitsch2013lasso} used stability selection to evaluate the significance of individual SNPs \citep{meinshausen2010stability}. \citet{guo2019combining} used a combination of both methods for feature selection, first selecting several $\lambda$ values via CV, and then using these penalty parameters for stability selection and final feature selection. As previously mentioned, however, the advisability of subsampling methods in the presence of correlated observations remains an open question \citep{roberts2017cross, jia2012preconditioning}. A rigorous assessment of penalty parameter selection in the context of penalized LMMs is beyond the scope of this work, but worthy of further investigation.}

@@

>>>>

@@

% As with multiple linear regression models, lasso penalized regression assume subjects are independent of each other with uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations.

As with multiple linear regression models, lasso penalized regression assume subjects are independent of each other with uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. The presence of structure can adversely affect analyses in several ways. Firstly, allele frequency differences due to ancestry within a sample can be mistaken for biological signal if steps are not taken to mitigate this effect. One common solution, dividing a sample and performing ancestry-specific analyses, often results in a loss of statistical power from reduced sample sizes. Secondly, correlation among subjects within a sample, whether due to common ancestry or more recent relatedness, violates an assumption that many regression methods rely on, rendering these statistical methods invalid.

Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, the majority of this work has focused on plant and animal breeding data, a setting which affords much greater experimental control than studies of human genetics \citep{Amin2007, hoffman2013correcting, price2006principal, rakitsch2013lasso, bhatnagar2020simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principle be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and confounding arising from observational study \citep{lawson2019population, barton2019population}.

@@

\begin{equation}
    Q(\bbeta | \lam) = \frac{1}{2n}|| \by -  \beta_0 - \bX \bbeta ||_2^2 + \lam ||\bbeta||_1,
\end{equation}
where $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||\mathbf{x}||_2 = \sqrt{\sum_{i=1}^p x_i^2}$ the $\ell_2$ norm. The first component of the objective function is the linear regression loss function, which quantifies the overall model fit and captures how well the estimated parameters agree with the observed data. As in ordinary least squares (OLS) regression, the lasso assumes observations are independent. The second component is the $\ell_1$ penalty applied to the estimated coefficient values (but not the intercept, $\beta_0$), which discourages large, unrealistic parameter values. The tuning parameter, $\lambda$, controls the tradeoff between these two components. For sufficiently large values of $\lambda$ all elements of $\bbeta$ will be estimated to be 0. In the $p \ll n$ setting, if $\lam = 0$, the lasso solution coincides with that of OLS. The $\ell_1$ penalty of the lasso can be applied to nonlinear models such as Cox regression \citep{tibshirani1997lasso} and logistic regression \citep{shevade2003simple} by replacing the residual sum of squares with the appropriate negative log-likelihood function.

@@

A central issue explored in this work has been the potential for population stratification and environmental confounding to lead to the identification of spurious associations. In Section~\ref{sec:sel_bias} we characterized the three lasso methods (ordinary, PC and LMM) in terms of their susceptibility to this type of selection bias in a simplified setting of two subpopulations. We now examine these methods' performance in a genome-wide data set with greater structural complexity.

Figure \ref{fig:gw_pc1_corr} displays the mean absolute correlation of the non-causal SNPs selected for a particular $\lambda$ value with the first principal component (PC1) of the Empirical data for each method across the full $\lambda$ path. The corresponding shaded regions indicate the SE about the mean. Overall, these bounds are largest early on in each $\lambda$ path, where only a few SNPs are be selected, but narrow as $\lambda$ increases and more variables are allowed to enter the model.\anna{Do I need to explain why we are using correlation with the first PC as a proxy for MAF difference or is this obvious?} The absolute correlation with PC1 is used as a measure of how population-informative the selected SNPs are given the complex Empirical data structure compared to that in Section~\ref{sec:sel_bias}.

@@

\anna{\textbf{Additional genome-wide results based on other figures we decide to include.} The results of a genome-wide analysis of Empirical data agree with the findings of our simulation experiments, in particular that PC-lasso and LMM-lasso are able to substantially reduce selection bias compared to the ordinary lasso. Although PC-lasso was found to control this selection bias most effectively, it should be noted that because the Empirical data consists of three distinct ethnic groups, its RRM (Figure~\ref{fig:empirical}) demonstrates predominantly coarse population structure. Given our results in Section~\ref{sec:sim-fine-coarse}, we might expect LMM-lasso to provide greater improvement over the ordinary lasso in the analysis of more finely structured genome-wide data.}

@@

\anna{The nature of the simulations and analyses in this work have allowed us to select penalty parameters based on the true number of causal features in a model, which allows for fair comparison across methods. In a real data scenario, the true number of causal SNPs would not be known a priori. As such, appropriate methods for penalty parameter selection in penalized LMMs are worth additional discussion. Penalized LMMs are a special case of what is more broadly known in the statistical literature as a preconditioned lasso algorithm, which refers to the pre-multiplication of $\bX$ and $\by$ by some matrix prior to running the ordinary lasso. At present the theoretical understanding of these algorithms remains fairly poor \citep{wauthier2013comparative}, and there is no clear consensus on the optimal way to select a penalty parameter for such algorithms.}

\anna{Several authors have used the Bayesian information criterion (BIC) for variable selection in mixed models \citep{bondell2010joint, schelldorfer2011estimation}. However it has been shown that the Akaike information criterion (AIC) and BIC may perform poorly in the high-dimensional setting \citep{fan2013tuning}. The high-dimensional BIC introduced by \citet{fan2013tuning} penalizes model parameters more severely in an attempt to perform better in high dimensional settings. \citet{bhatnagar2020simultaneous} employs this high-dimensional BIC as its default method to select $\lambda$.}

\anna{Subsampling approaches, such as CV and stability selection are another common class of variable selection methods often preferred to information criterion-based approaches in practice. \citet{rakitsch2013lasso} used stability selection to evaluate the significance of individual SNPs \citep{meinshausen2010stability}. \citet{guo2019combining} used a combination of both methods for feature selection, first selecting several $\lambda$ values via CV, and then using these penalty parameters for stability selection and final feature selection. As previously mentioned, however, the advisability of subsampling methods in the presence of correlated observations remains an open question \citep{roberts2017cross, jia2012preconditioning}. A rigorous assessment of penalty parameter selection in the context of penalized LMMs is beyond the scope of this work, but worthy of further investigation.}
